---
title: "SciViews HTTP server"
author: "Philippe Grosjean (phgrosjean@sciviews.org)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{SciViews HTTP server}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Principle

The {svHttp} package uses an option of the integrated HTML help system in R to develop a simple inter-process communication using URLs. It is designed to be the simplest as possible. Yet, it allows also to customize the way the server evaluates R code and sends results back to the client. An {svHttp} server with all default arguments is started in an R process using:

```{r}
library(svHttp)
server_port <- start_http_server()
server_port
```

The default port is 8888, but it can be changed with the `port =` argument in `start_http_server()` (the name of the server can also be changes with `name =` and modified or retrieved with `http_server_name()`).

Since there is only one HTML help system in R, starting the {svHttp} server most probably asks for a different port that the one previously chosen by R. In this case the R HTML help system is forced to close and to reopen with the requested port. There is no consequences for help pages displayed later on. A minor annoyance is in R \>= 4.2.0 where preexisting HTML help pages may not respond any more. In this case, just close and display the help page again to restore its full features.

On the client side on the same machine, a simple HTTP request can be used to execute R code. It only works locally to avoid security issues because the communication is simple and not encrypted (the HTTPS protocol is **not** available). The URL is `localhost:port/custom/SciViews` with an argument being the URL encoded R code to execute.

For instance to execute `2 - 1`, you could use the URL `http://localhost:8888/custom/SciViews?2-1`. With the {svHttp} server started on the default port 8888 in R, if you use this URL in any local web browser, you will get a page with `[1] -1`. This is the output R produces when the same instruction `2 - 1` is issued at the R console. Within a separate terminal, you can also issue `curl 'http://localhost:8888/custom/SciViews?2-1'` and obtain the same result. **By default, the {svHttp} server produces the same, or very similar outputs to what would be issued at the R console if the same command was issued there.** This is useful to implement a remote R console inside another application.

To get the content of an R variable, just use its name as message, as usual at the R prompt. You can get the content of `server_port` with the URL `http://localhost:8888/custom/SciViews?server_port`.

# Implementing a client

In R you can use, for instance, the {curl} package on the client. Here is a function that captures output of an R command on the {svHttp} server (note that the server sometimes issues a 500 error code while it still processes the command correctly, so make sure to catch that case too).

```{r}
http_server_run <- function(cmd, port = 8888) {
  # cmd is a string containing the command to process. We have to URLencode it
  cmd <- utils::URLencode(cmd)
  url <- paste0("http://localhost:", port, "/custom/SciViews?", cmd)
  res <- curl::curl_fetch_memory(url)
  if (res$status_code %in% c(200, 500)) {# Should be OK
    rawToChar(res$content)
  } else if (res$status_code > 0) {
    stop("Error while executing ", url, ": error ", res$status_code)
  }
}
```

Now, you could run arbitrary R commands in the {svHttp} server and get back the console output it produces. **Note: do not use this function in the *same* R process where the {svHttp} server runs.** As it is blocking, you would end up in a deadlock situation!

```{r, eval=FALSE}
cat(http_server_run("R.version"))
cat(http_server_run("http_server_name()"))
res <- http_server_run("ls()")
# Now you could do whatever you want with res
res
```

## Callback argument

A second argument can be passed to the URL: the name of a callback function. That way, the client gets this information to process the result send by the {svHttp} server. The URL is then `http://localhost:port/custom/SciViews?code&callback`. Also, there are numerous ways to customize the behavior of the {svHttp} server, see `?par_http_server`. A complete client written in JavaScript is implemented here: <https://github.com/SciViews/sciviewsk/blob/master/content/js/socket.js>. You could get inspiration from there to develop your own client application.

# Utilities on the server-side

Several functions are provided to manage the {svHttp} server on the R process where it runs. You can get or change the port or the name of the server with dedicated functions:

```{r}
http_server_port()
http_server_name()
# Change the name
http_server_name("myHttpServer")
```

At anytime, you can get an idea of the clients that are currently connected to the {svHttp} server (it can serve multiple clients at the same time, but it will , of course, process commands one at a time and the other clients are put on hold). There is a special client named `"default"` which process all requests by clients that did not identified themselves properly.

```{r}
http_server_clients()
```

Identification of clients is mandatory to allow further customization of the {svHttp} server in a stateful way. Again, examine the code here (<https://github.com/SciViews/sciviewsk/blob/master/content/js/socket.js>) for an example of complete implementation of all the features the {svHttp} server provides.

Finally, it is possible to close the {svHttp} server, but keep in mind that you also close the internal R HTML help service at the same time. This would only affect currently displayed help pages, since R would reopen the server automatically at the next help request.

```{r}
stop_http_server()
```

The `stop_http_server()` accepts an argument `remove.clients`, which is `FALSE` by default. If set to `TRUE`, the stateful configuration of registered clients is also deleted.

# Alternate solutions

-   To create a web API with defined functions, you can use {plumber}, see <https://www.rplumber.io>. Where {svHttp} allows for a simple, generic connection to an R process locally through an URL (any R command can be issued), {plumber} provides a defined set of actions. This allows to get a more secure server by restricting commands that the client can issue in the R process. However, a little bit more work is required to define the {plumber} API. The {svHttp} server also aims to replicate the behavior of R at the command prompt through URL requests, which is not the purpose of {plumber}.

-   The {svSocket} package (see <https://www.sciviews.org/svSocket/>) follows a very similar approach to {svHttp} but using sockets. This allows for more flexibility, and one has more control on the interactions between the client and the server. However, implementing an {svSocket} client is a little bit more complex.

-   The {Rserve} package provides a binary R server. When R objects are transferred from the R server to the client, they are binary converted into their Java counterparts. This could be useful to transfer large R objects to a client written in Java. However, both {svHttp} and {svSocket} are more suitable to implement a remote R console in an application written in any language.

-   The OpenCPU (<https://www.opencpu.org>) project aims to embed scientific computing in R into a web server. It gives a little bit more control on the way R code is executed and results are returned, but it is a little bit more complex to set up than {svHttp} that appears as more *ad hoc* solution for quick local inter-process computation in R.

There are several other server implementations and inter-process communication protocols implemented for R. See the CRAN task views, in particular, [web technologies and services](https://cran.r-project.org/view=WebTechnologies) and the [high-performance and parallel computing with R](https://cran.r-project.org/view=HighPerformanceComputing).
