---
title: "SciViews HTTP server"
author: "Philippe Grosjean (phgrosjean@sciviews.org)"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{SciViews HTTP server}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

TODO...

Use a HTTP request like this in the client: <http://127.0.0.1:8888/custom/SciViews?msg&callback>. The server returns something like (in a correct RJSONp object): `<callback>({"result":{"String 1", "String 2", "..."}, "options":"<options>", "name":"<server.name>","port":"<port>"})`. Another (simpler) way to call it is by using <http://127.0.0.1:8888/custom/SciViews?msg> and in this case, the client must manage the simple string returned.

The simplest test consists in starting the HTTP server in R:

```{r, eval=FALSE}
port <- start_http_server()
```

... and then, open a browser on the same machine and navigate to the following URL: <http://localhost:8888/custom/SciViews?port>. You should read there the port number as an answer (8888).
